Understanding Feedback Mechanisms in Machine Learning Jupyter Notebooks

### 2024-07-29 report/emse24.tex
feat: emse24 final draft

%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
% \documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
\usepackage[most]{tcolorbox}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{subcaption}
\usepackage{tabularx}
\usepackage{longtable}
\usepackage{booktabs}
\usepackage[colorinlistoftodos,prependcaption,textsize=tiny]{todonotes}
\usepackage{wrapfig}
\usepackage{framed}
\newcommand{\highlight}[1]{\begin{framed}%
  \noindent#1
\end{framed}}

@@

\title{Exploring the Role of Feedback in Machine Learning Jupyter Notebooks}

@@

\author{Arumoy Shome\and
  Lu{\`\i}s Cruz\and
  Diomidis Spinellis\and
  Arie van Deursen
}

@@

\begin{abstract}

  % TODO: update the abstract
  Developing and testing ML systems is a highly interactive process which demands a human-in-the-loop approach. In addition to writing tests for the code base, practitioners are required to analyse and interpret several visualisations using their domain expertise to validate if an ML system satisfies the required set of functional and non-functional properties. Visualisations are frequently used to qualitatively assess various parts of an ML pipeline. However, implicit knowledge gained from visualisations must be translated to explicit analytical tests that fail when there is change in any component of the ML pipeline. We conduct an empirical analysis of Jupyter notebooks to catalogue the state-of-the-art mappings between ML visualisations and assertions. We mine Github to collect 54K Jupyter Notebooks that contain assertions written in Python. We develop a novel methodology to identify 1764 notebooks which contain an assertion below a visualisation. We manually analyse the 1.7K notebooks and identify 269 visualisation-assertion pairs that are semantically related to one another. We further investigate the 269 visualisation-assertion pairs and identify three frequently occurring testing patterns. We perform an in-depth analysis of 34 visualisation-assertion pairs and find that the assertions often fail to capture all the information present in their visual counter-part. Empirical evidence obtained in this study indicates that current software testing methods fail to address the unique challenges of ML. And emphasises the need for automated tools that bridge the gap between visual assessments and analytical assertions.

\end{abstract}

\keywords{ML Testing \and SE4AI \and Visualisations \and Assertions \and Computational Notebooks}

\section{Introduction}
The machine learning development lifecycle (MLDL) encompasses a multifaceted series of entangled stages. Significant effort is invested in gathering and assembling the required data from diverse sources. The data then undergo cleaning and preprocessing to ensure high quality and usability, in the subsequent stages. MLDL is dominated by the model development phase which is highly experimental and iterative. This stage involves exploratory data analysis (EDA) to uncover underlying patterns, fitting the data to various models, and rigorously analysing the results. Often, this stage may lead to additional rounds of feature engineering, where new attributes are created based on insights gained from initial models to enhance model performance. Practitioners may also engage in comparing and contrasting different models to identify the most effective approach, followed by fine-tuning selected models to optimize performance. The final model is then deployed into production where it is continuously monitored, and periodically retrained to adapt to new data or changing conditions~\cite{haakman2021ai,amershi2019software,sculley2015hidden}.

@@

% NOTE: consider starting with prints and last statements first (most obvious), then move to assertions; its fine if RQs follow different order
Jupyter notebooks are primarily used to explore different ideas or approaches, often neglecting software development best practices in favor of rapid experimentation~\cite{kery2018story,rule2018exploration,pimentel2019large-scale}. During this process, practitioners can use various feedback mechanisms like assertions, print statements, output of code cells and visualisations to determine the appropriate data preprocessing techniques, detect erroneous assumptions, create new features to boost model performance, and select specific hyperparameters~\cite{head2019managing,chattopadhyay2020whats}. Once the code is finalised, it is typically transferred into a script that is run in a production environment. Alternatively, the notebook may contain significant exploratory work such as developing a model or performing data exploration. To effectively communicate the information present in the notebooks with clients, collaborators or teammates, the practitioners add annotations and structure to the notebook using markdown cells~\cite{kery2018story,rule2018exploration}. In both scenarios, practitioners need to revisit the notebooks to obtain structured information such as up-to-date and accurate requirements, their rationale and design decisions, and tests---artifacts which are usually missing in the original notebooks~\cite{pimentel2019large-scale,psallidas2019data,grotov2022large-scale}. As data, personnel, and system environments inevitably change, undocumented decisions can lead to accumulation of technical debt. This not only prolongs time required for debugging and troubleshooting, but also incurs additional costs and resource expenditures~\cite{sculley2015hidden,amershi2019software,sambasivan2021everyone}.

% TODO: feedback from diomidis
% - "We identify a knowledge gap in the current approach used to obtain such structured information from experimental code within Jupyter notebooks." Rephrase to have the knowledge gap correspond to what the RQs answer.  Currently one may think that we aim to improve the approach, not its understanding.
We identify a knowledge gap in the current approach used to obtain such structured information from experimental code within Jupyter notebooks. To address this gap, this study conducts a series of empirical studies to understand the role of various feedback mechanisms used to develop ML pipelines within Jupyter notebooks. We focus on implicit form of feedback obtained from print statements and output generated by the last statement in code cells, as well as explicit form of feedback obtained from assertions. To the best of our knowledge, we are the first to investigate how such feedback mechanisms are used by ML practitioners when developing ML pipelines inside Jupyter notebooks. The research questions along with the contributions of this study are presented below.

@@

    We mine 297.8 thousand Jupyter notebooks written in Python from Github and Kaggle. We analyse 2.3 million code cells obtained from the notebooks and create a public dataset of 27.1 thousand assertions, 498.3 thousand print statements and 373.6 thousand last statements. We perform a descriptive and lexical analysis of the dataset and report the key characteristics of feedback mechanisms used in ML Jupyter notebooks.

  % TODO: add citation for the dataset (make it public so that DOI is available)

@@

    To gain a deeper understanding of assert statements written in Jupter notebooks, we conduct individual case-studies with 82 assert statements. We analyse the assertion along with the surrounding code to understand its purpose. Additionally, we analyse the entire code cell, the previous and next cells along with the purpose of the notebook to bring in rich context.

  \item[RQ3.] \textbf{How is implicit feedback from print statements and last statements of code cells used when writing ML code in Jupyter notebooks?}

    Similarly, we conduct individual case-studies with 44 print statements and 27 last statements. During each case-study, we analyse the code statement along with the raw output to understand its purpose.
\end{description}

% TODO revisit this paragraph
% TODO what should be done to change the things we observe? to increase the use of explicit feedback? do we know if its important or not? maybe we need to explore that?
Our findings indicate that outputs from code cells are extensively utilised for making critical design and implementation decisions during the MLDL. While manual validation of these outputs is common, it is also prone to inconsistencies thus highlighting the need for automated validation mechanisms. Assertions are identified as a potential solution, capable of verifying data distributions and ensuring model performance metrics meet predefined thresholds, thereby reducing reliance on manual checks and enhancing the robustness of ML pipelines. However, the study also notes the limited use of assertions, with only 9.7\% of the 3 million notebooks analyzed incorporating them, suggesting a significant gap in current testing practices within the ML community. The findings underscore the necessity for developing ML-specific testing tools and integrating assertions more deeply into the ML development workflow to ensure consistent and reliable validation processes.

% TODO update replication package

% the replication package for this study is made public on figshare\footnote{the replication package can be found at this url: https://figshare.com/s/9347228011999cba29f5}.

@@

A Read-Eval-Print Loop (REPL) is an interactive programming environment that takes single user inputs (Read), executes them (Eval), returns the result to the user (Print) and then waits for the next command (Loop). REPLs offer an immediate feedback loop for quickly testing and debugging code snippets. The Python standard distribution includes a Python REPL with a command line interface to execute Python code in real-time. IPython---a more advanced alternative to the Python REPL---extends the capabilities of the standard Python shell with additional features such as powerful introspection, additional shell syntax, tab completion, and command history. Jupyter notebooks (and its successor Jupyter lab) leverage the IPython kernel to execute Python code and extends its capabilities with a graphical user interface (GUI). Resembling the literate programming paradigm~\cite{knuth1984literate}, Jupyter notebooks offer a web-based interface where users can create and share documents that contain live code, equations, visualisations, and narrative text. This format supports a literate approach to programming, making it easier to share and understand complex analyses and workflows. Jupyter Notebooks in particular, have become a popular tool in the ML community since they facilitate open and reproducible science, education, and data analysis.

@@

Figure~\ref{fig:nb} shows an example Jupyter notebook with markdown and code cells. The data comprised within Jupyter notebooks is stored using the JSON schema specified by \emph{nbformat}~\footnote{https://nbformat.readthedocs.io/en/latest/index.html}, and thus can be parsed programmatically. A notebook consists of \emph{cells} which can be of three types---``markdown'', ``code'' and ``raw''. All cells of a notebook are stored as a list under the top-level \lstinline[language={}]$cells$ key in the JSON representation. Within this list, each cell is represented as a JSON object. The contents of the cell are stored as a list of strings inside the \lstinline[language={}]$source$ key.

Only code cells in Jupyter notebooks can generate outputs. As shown in Figure~\ref{fig:nb}, outputs can come from three main sources in the source code---print statements, the last statement in the code cell, and statements that create a visualisation. The outputs produced by code cells are stored as JSON objects under the \lstinline[language={}]$outputs$ key in the corresponding JSON representation of the notebook. Text outputs (such as those produced by the last statement of a cell or print statements) are stored under the \lstinline[language={}]$text$ key. Visual outputs have a sightly different JSON schema and are stored under the \lstinline[language={}]$data$ key. The image itself is represented as a base64 encoded string which is stored inside the \lstinline[language={}]$data$ object, under the \lstinline[language={}]$image/png$ key.

@@

% TODO: I suspect this will raise eyebrows because the notebooks collected from Github were for the ICST paper, where we were also looking for the keyword "assert" (which does not apply to this paper)

Figure~\ref{fig:data-collection} illustrates the data collection process employed in this study to gather Jupyter notebooks written in Python from GitHub and Kaggle. We used Github's advanced search syntax~\footnote{https://docs.github.com/en/search-github/searching-on-github/searching-code}, with the following search query: \lstinline[language={}]$language:"Jupyter Notebook"$ to mine 49.9 thousand Jupyter notebooks from public repositories on Github~\footnote{Collected on June 22, 2023}. We used the pre-existing dataset KGTorrent~\cite{quaranta2021kgtorrent} for Kaggle since it does not support advanced code-based search like GitHub. To the best of our knowledge, KGTorrent is the largest dataset of Python Jupyter notebooks obtained from Kaggle, consisting of 247.9 thousand notebooks. Consequently, we started with 283GB of data comprising 297.8 thousand Python Jupyter notebooks.

Since the focus of this study is on the analysis of Python code, we only include notebooks that have a valid underlying JSON structure and contain at least one code cell. Finally, to focus on Python code written specifically for ML projects, we only include notebooks that import at least one of the following popular ML libraries which we identified from grey literature~\cite{datacampml,gfgml,jetbrainsml,kdnuggetml,courseraml}---Scikit Learn~\cite{virtanen2020scipy}, Pytorch~\cite{paszke2017automatic}, Tensorflow~\cite{abadi2015tensorflow} and Keras~\cite{chollet2015keras}. We use the Python \lstinline{ast} module to extract the assert, print and last statements from all code cells. Prior to conducting the analysis for RQ2 and RQ3, we remove duplicate data points resulting in a final data set of 27.1 thousand assertions, 498.3 thousand print statements and 373.6 thousand last statements.

@@

We allocated a fixed time resource of 200 hours to conduct the case study analysis of all candidate assertions, print statements and last statements. To surface interesting candidates for the case studies, we apply text processing techniques as described below.

The assertions and cell outputs are first tokenised---special characters and alpha-numeric words shorter than two characters are removed. Two stop words namely ``assert'' and ``print'' are removed since they appear in all assertions and print statements respectively. The term frequency (TF) for all tokens is calculated and then normalised using their inter-document frequency (IDF) such that tokens that appear less frequently are assigned a higher value. We apply stratified random sampling to identify the candidates for the case study analysis. The sub-groups are created by adding TF-IDF of the tokens in each candidate to produce an aggregate value. The candidates are then divided into quartiles based on the aggregate value. A candidate is randomly drawn from each bin and analysed as an in-depth case study. The analysis is stopped when the time resource is exhausted. During each case study, we analyse the code of the candidate to understand its purpose. Additionally, we analyse the entire code cell, the previous cell, next cell and the notebook's purpose to bring in rich context.

@@

% TODO: add citation to online appendix
The results of the study are presented in this section. For the case-studies conducted for RQ2 and RQ3, we group the candidates based on their similarity to each other. Please note that this is done only to facilitate structuring the paper in a meaningful manner and for easy of reporting the results. Figure~\ref{fig:taxonomy} provides an overview of the various feedback mechanism groups identified in this study. Additional details are provided in Section~\ref{sec:result-explicit} and Section~\ref{sec:result-implicit}. Throughout the text, we reference the candidates (\textbf{(A)}ssertions, \textbf{(P)}rint statements and \textbf{(L)}ast cell statements) using their unique identifiers as examples. These can be viewed on our online appendix.

@@

@@

Figure~\ref{fig:other-test-methods} presents the 10 most common methods used in the 5\% assertions written using external testing libraries. We manually tracked the source of the methods to the numpy, pandas, torch and unittest libraries. The most frequently used methods include \lstinline{assert_almost_equal} and \lstinline{assert_equal}, each with over 250 instances. Methods such as \lstinline{assert_almost_equal}, \lstinline{assert_array_almost_equal} and \lstinline{AssertAlmostEqual} accommodate the stochastic nature of ML, where results can vary slightly due to inherent randomness in data processing and model training. We also find methods such as \lstinline{assert_array_equal} and \lstinline{assert_frame_equal} that are specifically tailored for comparing high-dimensional data structures such as numpy arrays and pandas data frames.

\begin{figure}
  \centering
  \begin{minipage}{0.49\textwidth}
    \centering
    \includegraphics[width=\linewidth]{common-assert-test.pdf}
    \caption{10 most common tests used in \lstinline{assert} statements.}
    \label{fig:common-assert-test}
  \end{minipage}
  \hfill
  \begin{minipage}{0.49\textwidth}
    \centering
    \includegraphics[width=\linewidth]{common-compare-op.pdf}
    \caption{10 most common operators used in \lstinline{assert} statements that perform a comparison check.}
    \label{fig:common-compare-op}
  \end{minipage}
\end{figure}

\begin{table}
  \centering
  \caption{Examples of the most common tests performed in assert statements.}
  \begin{tabular}{@{}m{0.5\textwidth} m{0.5\textwidth}@{}}
    \toprule
    \emph{\textbf{Comparison Check}}&
    \emph{\textbf{Function Call}}\\
    \midrule

@@

We focus the remainder of the analysis on the 95\% of the assertions written using the \lstinline{assert} statement. The most common test performed in the \lstinline{assert} statements include a comparison check and a function call as shown in Figure~\ref{fig:common-assert-test}. Table~\ref{tab:common-assert-test} provides a few examples of assertions for each of the tests above. In Figure~\ref{fig:common-compare-op} we see the most common operators used in assert statements that perform a comparison check. We observe that overwhelming majority of the asserts ensure that two values are equal to each other. The shape and size of data structures are frequently compared as observed in Figure~\ref{fig:common-compare-lhs} and Figure~\ref{fig:common-compare-rhs} which show the most common bi-grams that appear in the LHS and RHS of the comparison statements respectively.

@@

@@

We collected 1.4 million print statements from 180 thousand notebooks (60\% of the total notebooks analysed in this study). Unlike assert statements, print statements do not have a fixed lexical semantic since the Python \lstinline{ast} module interprets them as regular function calls. Additionally, print statements can be written in several forms---the \lstinline{print} function may have multiple arguments, any form of string interpolation supported by Python or output from function calls. Figure~\ref{fig:common-print-nodes} shows that the most common AST nodes inside of print statements are constants followed by function calls. Figure~\ref{fig:common-print-constant-types} further validates that the constants are almost entirely of type string. Based on these observations, we disect the print statements into two components---text written in natural language (from the \emph{Constant} nodes) and code (from all other nodes). Figure~\ref{fig:common-print-constants} and Figure~\ref{fig:common-print-not-constants} present the most common bi-grams that appear in the natural text and code of print statements respectively. Both figures collectively indicate that print statements are frequently used to print various performance metrics of ML models.

\begin{figure}
  \centering
  \begin{minipage}{0.49\textwidth}
    \centering
    \includegraphics[width=\linewidth]{common-print-constants.pdf}
    \caption{Most common bi-grams that appear in natural text inside of print statements.}
    \label{fig:common-print-constants}
  \end{minipage}
  \hfill
  \begin{minipage}{0.49\textwidth}
    \centering
    \includegraphics[width=\linewidth]{common-print-not-constants.pdf}
    \caption{Most common bi-grams that appear in the code of print statements.}
    \label{fig:common-print-not-constants}
  \end{minipage}
\end{figure}

>>>>

%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
% \documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
\usepackage[most]{tcolorbox}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{subcaption}
\usepackage{tabularx}
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{wrapfig}
\usepackage{framed}
\newcommand{\highlight}[1]{\begin{framed}%
  \noindent#1
\end{framed}}

@@

\title{Understanding Feedback Mechanisms in Machine Learning Jupyter Notebooks}

@@

\author{Arumoy Shome\and
Lu\`{i}s Cruz\and
  Diomidis Spinellis\and
  Arie van Deursen
}

@@

\begin{abstract}
  The machine learning development lifecycle involves iterative and exploratory processes that heavily rely on feedback mechanisms to ensure data and model integrity. This study investigates the use of implicit and explicit feedback within Jupyter notebooks. We analyse 297.8 thousand Jupyter notebooks from GitHub and Kaggle, identifying key feedback mechanisms such as print statements, outputs from last cell statements, and assertions. Our findings reveal extensive use of implicit feedback for critical design decisions and highlight the limited adoption of explicit feedback mechanisms like assertions. By conducting detailed case studies on selected feedback instances, we uncover the reliance on manual validation and the potential for automated validation to enhance robustness and maintainability. This study underscores the need for improved documentation and testing practices within the ML community, advocating for the integration of feedback mechanisms into the ML development workflow to mitigate technical debt and enhance reproducibility.
\end{abstract}

\keywords{Assertions, Feedback Mechanisms, Jupyter notebooks, SE4AI}

\section{Introduction}

The machine learning development lifecycle (MLDL) encompasses a multifaceted series of entangled stages. Significant effort is invested in gathering and assembling the required data from diverse sources. The data then undergoes cleaning and preprocessing to ensure high quality and usability, in the subsequent stages. MLDL is dominated by the model development phase which is highly experimental and iterative. This stage involves exploratory data analysis (EDA) to uncover underlying patterns, fitting the data to various models, and rigorously analysing the results. Often, this stage may lead to additional rounds of feature engineering, where new attributes are created based on insights gained from initial models to enhance model performance. Practitioners may also engage in comparing and contrasting different models to identify the most effective approach, followed by fine-tuning selected models to optimize performance. The final model is then deployed into production where it is continuously monitored, and periodically retrained to adapt to new data or changing conditions~\cite{haakman2021ai,amershi2019software,sculley2015hidden}.

@@

Notebook users can obtain feedback using implicit or explicit mechanisms. Implicit feedback mechanisms require manual validation by the user. For instance, a common prerequisite for all ML models is that the features in the dataset are all numerical. This can be verified with a print statement in a code cell and manually checking the output. This code statement however, will continue to execute even if the data type of a feature changes in subsequent batches of training data.

In contrast, explicit feedback mechanisms document the expected conditions of the code and halt execution if these conditions are not met. For instance, the implicit validation method mentioned above can be replaced with an explicit assertion: \lstinline{assert all(df[col].dtype in ['int64', 'float64'] for col in df.columns), "Not all columns are of numerical type"}. This assert statement will stop the execution if the condition is not satisfied and provide immediate feedback with a clear message explaining the issue.

As machine learning pipelines developed within Jupyter notebooks become business-critical, they are frequently transitioned into automated scripts for production environments~\cite{kery2018story,rule2018exploration}. In these contexts, it is crucial to shift from implicit feedback mechanisms to explicit ones. This transition reduces the reliance on manual verification, thereby enhancing the robustness and maintainability of ML pipelines.

To the best of our knowledge, no prior studies have been conducted to understand how feedback mechanisms are used by practitioners when developing ML pipelines inside Jupyter notebooks. To address this gap, we conduct a series of empirical studies to understand the role of various feedback mechanisms within Jupyter notebooks. We focus on implicit form of feedback obtained from print statements and output generated by the last statement of code cells, as well as explicit form of feedback obtained from assertions. The research questions along with the contributions of this study are presented below.

@@

    We mine 297.8 thousand Jupyter notebooks written in Python from GitHub and Kaggle. We analyse 2.3 million code cells obtained from the notebooks and create a public dataset of 89.6 thousand assertions, 1.4 million print statements and 1 million last cell statements. We perform a descriptive and lexical analysis of the dataset and report the key characteristics of feedback mechanisms used in ML Jupyter notebooks. The dataset along with the replication package is shared publicly under the Creative Commons Attribution (CC-BY) licence~\cite{replication}.

@@

    To gain a deeper understanding of Python assertions written in Jupter notebooks, we conduct individual case-studies with 82 assert statements. We analyse the assertion along with the surrounding code to understand its purpose. Additionally, we analyse the entire code cell, the previous and next cells along with the purpose of the notebook to bring in rich context.

  \item[RQ3.] \textbf{How is implicit feedback from print statements and last cell statements used when writing ML code in Jupyter notebooks?}

    Similarly, we conduct individual case-studies with 44 print statements and 27 last cell statements. During each case-study, we analyse the code statement along with the raw output to understand its purpose.
\end{description}

% The above all motivates the research: but then the intro should also ¡®sell¡¯ the outcomes of the research. These are
% - Identification of over 20 feedback mechanisms, half of them implicit, half of them explicit (your taxonomy)
% - Insights in actual use of these mechanisms
% - Recommendations for better support of these mechanisms in notebook environments / life cycle

Our findings indicate that implicit feedback mechanisms are extensively utilized for making critical design and implementation decisions during the MLDL. While manual validation of these outputs is common, it is also prone to inconsistencies thus highlighting the need for automated validation mechanisms. Assertions are identified as a potential solution, capable of verifying data distributions and ensuring model performance metrics meet predefined thresholds. However, the study also notes the limited use of assertions, with only 8.5\% of the notebooks analysed incorporating them, suggesting a significant gap in current testing practices within the ML community. The findings underscore the necessity for developing ML-specific testing tools that are deeply integrated into the ML development workflow to ensure consistent and reliable validation processes.

@@

A Read-Eval-Print Loop (REPL) is an interactive programming environment that takes single user inputs (Read), executes them (Eval), returns the result to the user (Print) and then waits for the next command (Loop). REPLs offer an immediate feedback loop for quickly testing and debugging code snippets. The Python standard distribution includes a Python REPL with a command line interface to execute Python code in real-time. IPython---a more advanced alternative to the Python REPL---extends the capabilities of the standard Python shell with additional features such as powerful introspection, additional shell syntax, tab completion, and command history. Jupyter notebooks (and its successor Jupyter lab) leverage the IPython kernel to execute Python code and extends its capabilities with a graphical user interface (GUI). Resembling the literate programming paradigm~\cite{knuth1984literate}, Jupyter notebooks offer a web-based interface where users can create and share documents that contain live code, equations, visualizations, and narrative text. This format supports a literate approach to programming, making it easier to share and understand complex analyses and workflows. Jupyter Notebooks in particular, have become a popular tool in the ML community since they facilitate open and reproducible science, education, and data analysis.

@@

Figure~\ref{fig:nb} shows an example Jupyter notebook with markdown and code cells. The data comprised within Jupyter notebooks is stored using the JSON schema specified by \emph{nbformat}\footnote{https://nbformat.readthedocs.io/en/latest/index.html}, and thus can be parsed programmatically. A notebook consists of \emph{cells} which can be of three types---``markdown'', ``code'' and ``raw''. All cells of a notebook are stored as a list under the top-level \lstinline[language={}]$cells$ key in the JSON representation. Within this list, each cell is represented as a JSON object. The contents of the cell are stored as a list of strings inside the \lstinline[language={}]$source$ key.

Only code cells in Jupyter notebooks can generate outputs. As shown in Figure~\ref{fig:nb}, outputs can come from three main sources in the source code---print statements, the last statement in the code cell, and statements that create a visualization. The outputs produced by code cells are stored as JSON objects under the \lstinline[language={}]$outputs$ key in the corresponding JSON representation of the notebook. Text outputs (such as those produced by the last statement of a cell or print statements) are stored under the \lstinline[language={}]$text$ key. Visual outputs have a sightly different JSON schema and are stored under the \lstinline[language={}]$data$ key. The image itself is represented as a base64 encoded string which is stored inside the \lstinline[language={}]$data$ object, under the \lstinline[language={}]$image/png$ key.

@@

Figure~\ref{fig:data-collection} illustrates the data collection process employed in this study to gather Jupyter notebooks written in Python from GitHub and Kaggle. We used GitHub's advanced search syntax\footnote{https://docs.github.com/en/search-github/searching-on-github/searching-code}, with the following search query: \lstinline[language={}]$language:"Jupyter Notebook"$ to mine 49.9 thousand Jupyter notebooks from public repositories on GitHub\footnote{Collected on June 22, 2023}. We used the pre-existing dataset KGTorrent~\cite{quaranta2021kgtorrent} for Kaggle since it does not support advanced code-based search like GitHub. To the best of our knowledge, KGTorrent is the largest dataset of Python Jupyter notebooks obtained from Kaggle, consisting of 247.9 thousand notebooks. Consequently, we started with 283 GB of data comprising 297.8 thousand Python Jupyter notebooks.

Since the focus of this study is on the analysis of Python code, we only include notebooks that have a valid underlying JSON structure and contain at least one code cell. Finally, to focus on Python code written specifically for ML projects, we only include notebooks that import at least one of the following popular ML libraries---Scikit Learn~\cite{virtanen2020scipy}, Pytorch~\cite{paszke2017automatic}, Tensorflow~\cite{abadi2015tensorflow} and Keras~\cite{chollet2015keras}. The libraries are selected based on the import analysis conducted by \emph{Psallidas et al.}~\cite{psallidas2019data} on 6 million Jupyter notebooks collected from GitHub. We use the Python \lstinline{ast} module to extract the assert, print and last statements from all code cells. Prior to conducting the analysis for RQ2 and RQ3, we remove duplicate data points resulting in a data set of 27.1 thousand assertions, 498.3 thousand print statements and 373.6 thousand last cell statements.

@@

For RQ2 and RQ3, we allocated a fixed time resource of 200 hours to conduct the case study analysis of all candidate assertions, print statements and last cell statements. To surface interesting candidates for the case studies, we apply text processing techniques as described below.

The assertions, print statements and last cell statements are first tokenized---special characters and alphanumeric words shorter than two characters are removed. Two stop words namely ``assert'' and ``print'' are removed since they appear in all assertions and print statements respectively. The term frequency (TF) for all tokens is calculated and then normalized using their inter-document frequency (IDF) such that tokens that appear less frequently are assigned a higher value. We apply stratified random sampling to identify the candidates for the case study analysis. The subgroups are created by adding TF-IDF of the tokens in each candidate to produce an aggregate value. The candidates are then divided into quartiles based on the aggregate value. A candidate is randomly drawn from each bin and analysed as an in-depth case study. The analysis is stopped when the time resource is exhausted.

During each case study, we analyse the code of the candidate to understand its purpose. Additionally, we analyse the entire code cell, the previous cell, next cell and the notebook's purpose to bring in rich context. For case studies where the first author was unable to determine the purpose of the candidate, it was additionally analysed by the second and third authors. Further discussions were held with all authors until a consensus was reached.

@@

% TODO: add citation to online appendix

The results of the study are presented in this section. For the case studies conducted for RQ2 and RQ3, we group the candidates based on their similarity to each other. This is done only to structure the paper in a meaningful manner and for ease of reporting the results. Figure~\ref{fig:taxonomy} provides an overview of the various feedback mechanism groups identified in this study with additional details provided in Section~\ref{sec:result-explicit} and Section~\ref{sec:result-implicit}. Throughout the text, we reference the candidates (\textbf{(A)}ssertions, \textbf{(P)}rint statements and \textbf{(L)}ast cell statements) using their unique identifiers as examples. These can be viewed on our online appendix.

@@

% TODO: add the findings from this part of the analysis

@@

Figure~\ref{fig:other-test-methods} presents the 10 most common methods used in the 5\% assertions written using external testing libraries. We manually tracked the source of the methods to the numpy, pandas, torch and unittest libraries. The most frequently used methods include \lstinline{assert_almost_equal} and \lstinline{assert_equal}, each with over 250 instances. Methods such as \lstinline{assert_almost_equal}, \lstinline{assert_array_almost_equal} and \lstinline{AssertAlmostEqual} accommodate the stochastic nature of ML, where results can vary slightly due to inherent randomness in data processing and model training. We also find methods such as \lstinline{assert_array_equal} and \lstinline{assert_frame_equal} that are specifically tailored for comparing high-dimensional data structures such as numpy arrays and pandas dataframes.

\highlight{\textbf{Finding 1.1} 25 thousand (8.5\%) notebooks contained at least one assertion. 95\% of the 89.6 thousand assertions were written using the built-in Python \lstinline{assert} statement and 5\% using external testing libraries. Methods such as \lstinline{assert_almost_equal} and \lstinline{assert_array_equal} from the numpy and pandas libraries were frequently used to accomodate for the stochastic nature of ML algorithms and comparing high-dimensional data structures.}

\begin{figure}
  \centering
  \begin{minipage}{0.49\textwidth}
    \centering
    \includegraphics[width=\linewidth]{common-assert-test.pdf}
    \caption{10 most common AST nodes in the test attribute of \lstinline{assert} statements.}
    \label{fig:common-assert-test}
  \end{minipage}
  \hfill
  \begin{minipage}{0.49\textwidth}
    \centering
    \includegraphics[width=\linewidth]{common-compare-op.pdf}
    \caption{10 most common operators used in \lstinline{assert} statements that perform a comparison check.}
    \label{fig:common-compare-op}
  \end{minipage}
\end{figure}

\begin{table}
  \centering
  \caption{Examples of the most common tests performed in \lstinline{assert} statements.}
  \begin{tabular}{@{}m{0.5\textwidth} m{0.5\textwidth}@{}}
    \toprule
    \emph{\textbf{Comparison Check}}&
    \emph{\textbf{Function Call}}\\
    \midrule

@@

We focus the remainder of the analysis on the 95\% of the assertions written using the \lstinline{assert} statement. Figure~\ref{fig:common-assert-test} shows the most frequent AST nodes that occur in \lstinline{assert} statements. The figure indicates that the most common test performed in the \lstinline{assert} statements include a comparison check and a function call. Table~\ref{tab:common-assert-test} provides a few examples of assertions for each of the tests above. In Figure~\ref{fig:common-compare-op} we see the most common operators used in assert statements that perform a comparison check. We observe that overwhelming majority of the asserts ensure that two values are equal to each other. Figure~\ref{fig:common-compare-lhs} and Figure~\ref{fig:common-compare-rhs} show the most common bi-grams that appear in the LHS and RHS of the comparison statements respectively. The figures collectively show that the shape and size of data structures are frequently checked.

@@

\highlight{\textbf{Finding 1.2} 31.7\% of the assertions written using the built-in \lstinline{assert} statement have a failure message. These assertions frequently perform a comparison check or call an external function. The comparison checks frequently verify that the shape and size of data structures is equal to a specified value.}

@@

We collected 1.4 million print statements from 180 thousand notebooks (60\% of the total notebooks analysed in this study). Unlike assert statements, print statements do not have a fixed lexical semantic since the Python \lstinline{ast} module interprets them as regular function calls. Additionally, print statements can be written in several forms---the \lstinline{print} function may have multiple arguments, any form of string interpolation supported by Python or output from function calls. Figure~\ref{fig:common-print-nodes} shows that the most common AST nodes inside of print statements are constants followed by function calls. Figure~\ref{fig:common-print-constant-types} further validates that the constants are almost entirely of type string. Based on these observations, we dissect the print statements into two components---text written in natural language (from the \emph{Constant} nodes) and code (from all other nodes). Figure~\ref{fig:common-print-constants} and Figure~\ref{fig:common-print-not-constants} present the most common bi-grams that appear in the natural text and code of print statements respectively. Both figures collectively indicate that print statements are frequently used to print various performance metrics of ML models.

\highlight{\textbf{Finding 1.3} Print statements are most commonly used to display various performance metrics of ML models.}

\begin{figure}
  \centering
  \begin{minipage}{0.49\textwidth}
    \centering
    \includegraphics[width=\linewidth]{common-print-constants.pdf}
    \caption{Most common bi-grams that appear in natural text of print statements.}
    \label{fig:common-print-constants}
  \end{minipage}
  \hfill
  \begin{minipage}{0.49\textwidth}
    \centering
    \includegraphics[width=\linewidth]{common-print-not-constants.pdf}
    \caption{Most common bi-grams that appear in the code of print statements.}
    \label{fig:common-print-not-constants}
  \end{minipage}
\end{figure}

We collected 1 million last cell statements from 119 thousand notebooks (40\% of the total notebooks analysed in this study). Figure~\ref{fig:common-last-nodes} shows that \emph{Call} nodes frequently occur in the AST of last cell statements. Figure~\ref{fig:common-last-modules} shows the most common Python modules and objects while Figure~\ref{fig:common-last-functions} show the most common functions used inside these \emph{Call} nodes. The figures highlight the prominence of data visualization libraries seaborn (\lstinline{sns}) and matplotlib (\lstinline{plt}) and visualization functions such as \lstinline{plot} and \lstinline{countplot} in the notebooks. The data analysis tool pandas (represented by \lstinline{df} and \lstinline{pd}) also features prominently with many data exploration and preprocessing functions such as \lstinline{head}, \lstinline{value_counts} and \lstinline{describe} being used frequently. The presence of \lstinline{train} and \lstinline{model} Python objects along with the \lstinline{fit} function indicates the use of ML libraries.

\highlight{\textbf{Finding 1.4} Libraries and functions for data visualization, data exploration and machine learning are most common in last cell statements.}

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{common-last-nodes.pdf}
  \caption{Most common AST nodes in the last cell statements.}
  \label{fig:common-last-nodes}
\end{figure}
